---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Setting global knitr parameters to print 2 decimals.

```{r echo = TRUE}
library(knitr)
options(scipen=1, digits=2)
```

## Loading and preprocessing the data

```{r echo = TRUE}
library(dplyr)
library(ggplot2)
act <- read.csv('activity.csv')
act$date <- as.Date(act$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Grouping the data by day and summarizing using the sum function:

```{r}
act_grouped_day <- group_by(act, date)
act_steps_day <- summarise(act_grouped_day, steps_per_day = sum(steps, na.rm = FALSE))
```

### Histogram of number of steps per day.

```{r}
qplot(act_steps_day$steps_per_day, fill=I('darkred'), col=I('black'), main='Histogram of total number of steps per day', xlab='Number of steps per day', ylab = 'Frequency')
```

The mean number of steps per day is `r mean(act_steps_day$steps_per_day, na.rm = TRUE)` and the median is `r median(act_steps_day$steps_per_day, na.rm = TRUE)`

## What is the average daily activity pattern?

Grouping the data by interval and summarizing using the mean function:

```{r echo = TRUE}
act_interval_group <- group_by(act, interval)
act_interval_avg <- summarise(act_interval_group, average_steps = mean(steps, na.rm = TRUE))
with(act_interval_avg, qplot(interval, average_steps, col=I('darkred'), geom = "line", xlab = "5-minute interval", ylab = "Average number of steps per day", main = "Time series of the 5-minute interval and its average number of steps."))
```

The 5-minute interval with the maximum average number of steps per day is `r act_interval_avg[which(act_interval_avg$average_steps == max(act_interval_avg$average_steps)), ]['interval']`

## Imputing missing values

There are `r sum(is.na(act$steps))` missing values in the dataset, which represents `r 100 * sum(is.na(act$steps)) / dim(act)[1]`% of the data.

I will use a straightforward method to impute the missing values, using the average 5-minute interval of each day.

```{r echo = TRUE}
act_na_filled <- act
act_na_filled[act_na_filled$interval == act_interval_avg$interval & is.na(act_na_filled), ]$steps <- act_interval_avg$average_steps
```

```{r echo = TRUE}
nafilled_grouped_day <- group_by(act_na_filled, date)
nafilled_steps_day <- summarise(nafilled_grouped_day, steps_per_day = sum(steps, na.rm = FALSE))
qplot(nafilled_steps_day$steps_per_day, fill=I('darkred'), col=I('black'), main='Histogram of total number of steps per day', xlab='Number of steps per day', ylab = 'Frequency')
```

The mean number of steps per day with imputed missing values is `r mean(nafilled_steps_day$steps_per_day, na.rm = TRUE)` and the median is `r median(nafilled_steps_day$steps_per_day, na.rm = TRUE)`, almost the same as with the missing values.

## Are there differences in activity patterns between weekdays and weekends?

```{r echo = TRUE}
act_na_filled$week_day <- sapply(act_na_filled$date, FUN = function(x) {if (weekdays(x) == 'Sunday' | weekdays(x) == 'Saturday') {'Weekend'} else {'Weekday'}})
act_na_filled$week_day <- factor(act_na_filled$week_day)
act_na_filled_interval_group <- group_by(act_na_filled, interval, week_day)
act_na_filled_int <- summarise(act_na_filled_interval_group, average_steps = mean(steps, na.rm = FALSE))
```

Panel plot comparing the time series of the 5-minute interval between weekdays and wekkends.

```{r echo = TRUE}
with(act_na_filled_int, qplot(interval, average_steps, facets=week_day ~ ., geom = "line", xlab = "5-minute interval",col=I('darkred'), ylab = "Average number of steps per day", main = "Time series of the 5-minute interval and its average number of steps per week day."))
```

The maximum number of steps is lower at the weekends.